### 2016-05-16 text/Chapter05.tex
Spell checking

The \emph{flat coeffect calculus} introduced in the previous chapter uniformly captures a number
of context-aware systems outlined in Chapter~\ref{ch:applications}. The coeffect calculus
can be seen as a \emph{language framework} that simplifies the construction of concrete
\emph{domain-specific} coeffect lnguages. In the previous chapter, we discussed how it provides a
type system that tracks the required context. In this chapter, we show that the language framework
also provides a way for defining the semantics of concrete domain-specific coeffect languages,
guides their implementation and simplifies safety proofs.

@@

This chapter links together a number of different technical developments presented in this thesis.
We take the flat coeffect calculus introduced in Chatper~\ref{ch:flat}, define its \emph{abstract
comonadic semantics} and use it to define a translation that gives a \emph{concrete operational
semantics} to a number of concrete context-aware languages. The type system is used to guarantee
that the resulting programs are correct. Finally, the development in this chapter is closely
mirrored by the implementation presented in Chapter~\ref{ch:impl}, which implements the
translation together with an interpreter for the target language.

@@

For a well-defined semantics, we need to ensure that a well-typed term is assigned exactly one
meaning. This can be achieved in a number of ways. First, we can prove the \emph{coherence}
\cite{semantics-bltres} and show the morhphisms assigned to multiple typing derivations are equivalent.
Second, the typing judgement can have a unique typing derivation. We follow the latter approach,
using the uniqye typing derivation specified in Section~\ref{sec:flat-unique}.

@@

The monadic and comonadic model outlined aboe represents at most a binary analysis of effects or
context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no
context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context\footnote{This is an
over-simplification as we can use \eg~stacks of monad transformers and model functions with
two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers
require the user to define complex systems of lifting to be composable. Consequently, they are usually used
for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
capturing fine-grained properties (\eg~a set of memory regions that may be accessed by a
stateful computation).}.

@@

\paragraph{Semantics of subcoeffecting.}
Although we do not include subcoeffecting in the core flat coeffect calculus, it is an interesting
extension to consider. Semantically, subcoeffecting drops some of the available contextual
capabilities (drops some of the implicit parameters or some of the past values). This can be
modelled by adding a (family of) lifting operation(s):
%
\begin{itemize}
 \item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
   $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
\end{itemize}
%
The axioms of flat coeffect algebra do not, in general, require that
$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$,
but the property holds for the three sample coeffect systems we consier. For systems with
the above property, the \ident{split} operation can be expressed in terms of lifting
(subcoeffecting) as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c),~
    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
\end{array}
\end{equation*}
%
The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object
mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map}
for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
are first and second projections from a two-element pair. To define the
$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
to throw away additional parts of the context and then transform the values in the
context.

@@

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{prj}): $e = \pi_i(v_1, \ldots, v_n)$ and so the last applied typing rule must have been
  (\emph{tup}) and $\Gamma \vdash (v_1, \ldots, v_n) : \tau_1 \times\ldots\times \tau_n$ and
  $\tau = \tau_i$. After applyng (\emph{prj}) reduction, $e' = v_i$ and so $\Gamma \vdash e' : \tau_i$.

@@

\paragraph{Expressions.}
The rules shown in Figure~\ref{fig:semantics-translation} define how expressions of the source
language are translated into the target language. The rules are very similar to those shown earlier
in Figure~\ref{fig:semantics-flat}. The consequent is now written as source code in the
target programming language rather than as composition of morphisms in a category. However,
thanks to the relationship between $\lambda$-calculus and cartesian closed categories, both
interpretations are equivalent.

@@

\begin{theorem}[Well-typedness of the translation]
\label{thm:semantics-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$
written in a context-aware programming langugae that is translated to the target language as
(we write $\ldots$ for the omitted part of the translation tree):
%
\begin{equation*}
  \semdeff
    {~~(\ldots)~~}
    {\coctx{}{\cclrd{r}} \vdash e : \tau }
    {(\ldots)}
    { ~~f }
\end{equation*}
%
Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation. Given a judgement
$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\cclrd{c}} \vdash e : \tau$, the translation
constructs a function of type $\ctyp{\cclrd{c}}(\sem{\tau_1} \times\ldots\times \sem{\tau_n}) \rightarrow \sem{\tau}$.

@@

The Figure~\ref{fig:semantics-ext-df} extensionds the target language with syntax, typing rules,
additional translation rule and reductions for modelling dataflow computations. We introduce a
new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
and a matching kind of expressions. We specify how the \kvd{prev} keyword is translated into a
$\ident{prev}_\cclrd{r}$ operation of the target language and we also add a typing rule
(\emph{df}) that checks the types of the elements of the stream and also guarantees
that the number of elements in the stream matches the number in the coeffect.
The additional reduction rules mirror the semantics that we discussed in
Example~\ref{thm:semantics-indexed-list} when discussing the indexed list comonad.

@@

\paragraph{Properties.}
Now consider a target language consisting of the core (ML-subset) defined by the syntax,
reduction rules and typing rules given in Figure~\ref{fig:semantics-target} and comonadically-inspired
primtives defined in Figure~\ref{fig:semantics-ext} and also concrete notion of comonadically-inspired
value and reduction rules for dataflow as defined in Figure~\ref{fig:semantics-ext-df}.

@@

\paragraph{Parameteric effect monads.}
Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our
indexed comonads) are closely related to our definition.  Although presented in a more general
categorical framework (and using monads), the model (i) defines the \ident{unit} operation only on the
unit of a monoid and (ii) the \ident{bind} operation composes effect annotations using the provided
monoidal structure.

@@

\begin{itemize}
\item Semantically, coeffects capture different notions of computation. As demonstrated in
  Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
  computation, many of which cannot be captured by a monad (\eg~liveness or dataflow).
  In terms of program analysis \cite{other-optcomp}, monas capture forward dataflow analyses and
  comonads correspond to backward dataflow analyses.

@@

\noindent
Despite the differences, our implicit paremeters resemble, in many ways, the \emph{reader} monad.
As discussed in Section~\ref{sec:semantics-related-monads}, the \emph{reader} monad is semantically
equivalent to the \emph{product} comonad when we consider just sequential composition. For a
language with lambda abstraction, we need a slight extension to the usual treatment of monads in
order to model implicit parameters using a monad.

@@

\noindent
This equivalence shows an intriguing relationship between the \emph{product comonad} and
\emph{reader monad}, but it cannot be extended beyond that. In particular, comonads that model
dataflow computations or liveness do not have a corresponding monadic structure.
This equivalence holds for monads and comonads (as well as \emph{indexed} monads
and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging. This can be
supported in monadic computations by adding an aditional operation discussed next.

>>>>

The \emph{flat coeffect calculus} introduced in the previous chapter uniformly captures a number
of context-aware systems outlined in Chapter~\ref{ch:applications}. The coeffect calculus
can be seen as a \emph{language framework} that simplifies the construction of concrete
\emph{domain-specific} coeffect languages. In the previous chapter, we discussed how it provides a
type system that tracks the required context. In this chapter, we show that the language framework
also provides a way for defining the semantics of concrete domain-specific coeffect languages,
guides their implementation and simplifies safety proofs.

@@

This chapter links together a number of different technical developments presented in this thesis.
We take the flat coeffect calculus introduced in Chapter~\ref{ch:flat}, define its \emph{abstract
comonadic semantics} and use it to define a translation that gives a \emph{concrete operational
semantics} to a number of concrete context-aware languages. The type system is used to guarantee
that the resulting programs are correct. Finally, the development in this chapter is closely
mirrored by the implementation presented in Chapter~\ref{ch:impl}, which implements the
translation together with an interpreter for the target language.

@@

For a well-defined semantics, we need to ensure that a well-typed term is assigned exactly one
meaning. This can be achieved in a number of ways. First, we can prove the \emph{coherence}
\cite{semantics-bltres} and show the morhphisms assigned to multiple typing derivations are equivalent.
Second, the typing judgement can have a unique typing derivation. We follow the latter approach,
using the unique typing derivation specified in Section~\ref{sec:flat-unique}.

@@

The monadic and comonadic model outlined above represents at most a binary analysis of effects or
context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no
context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context\footnote{This is an
over-simplification as we can use \eg~stacks of monad transformers and model functions with
two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers
require the user to define complex systems of lifting to be composable. Consequently, they are usually used
for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
capturing fine-grained properties (\eg~a set of memory regions that may be accessed by a
stateful computation).}.

@@

\paragraph{Semantics of subcoeffecting.}
Although we do not include subcoeffecting in the core flat coeffect calculus, it is an interesting
extension to consider. Semantically, subcoeffecting drops some of the available contextual
capabilities (drops some of the implicit parameters or some of the past values). This can be
modelled by adding a (family of) lifting operation(s):
%
\begin{itemize}
 \item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
   $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
\end{itemize}
%
The axioms of flat coeffect algebra do not, in general, require that
$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$,
but the property holds for the three sample coeffect systems we consider. For systems with
the above property, the \ident{split} operation can be expressed in terms of lifting
(subcoeffecting) as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c),~
    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
\end{array}
\end{equation*}
%
The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object
mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map}
for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
are first and second projections from a two-element pair. To define the
$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
to throw away additional parts of the context and then transform the values in the
context.

@@

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{prj}): $e = \pi_i(v_1, \ldots, v_n)$ and so the last applied typing rule must have been
  (\emph{tup}) and $\Gamma \vdash (v_1, \ldots, v_n) : \tau_1 \times\ldots\times \tau_n$ and
  $\tau = \tau_i$. After applying (\emph{prj}) reduction, $e' = v_i$ and so $\Gamma \vdash e' : \tau_i$.

@@

\paragraph{Expressions.}
The rules shown in Figure~\ref{fig:semantics-translation} define how expressions of the source
language are translated into the target language. The rules are very similar to those shown earlier
in Figure~\ref{fig:semantics-flat}. The consequent is now written as source code in the
target programming language rather than as composition of morphisms in a category. However,
thanks to the relationship between $\lambda$-calculus and Cartesian closed categories, both
interpretations are equivalent.

@@

\begin{theorem}[Well-typedness of the translation]
\label{thm:semantics-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\cclrd{r}} \vdash e : \tau$
written in a context-aware programming language that is translated to the target language as
(we write $\ldots$ for the omitted part of the translation tree):
%
\begin{equation*}
  \semdeff
    {~~(\ldots)~~}
    {\coctx{}{\cclrd{r}} \vdash e : \tau }
    {(\ldots)}
    { ~~f }
\end{equation*}
%
Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{\Gamma}{\cclrd{r}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation. Given a judgement
$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\cclrd{c}} \vdash e : \tau$, the translation
constructs a function of type $\ctyp{\cclrd{c}}(\sem{\tau_1} \times\ldots\times \sem{\tau_n}) \rightarrow \sem{\tau}$.

@@

The Figure~\ref{fig:semantics-ext-df} extends the target language with syntax, typing rules,
additional translation rule and reductions for modelling dataflow computations. We introduce a
new kind of values written as $\ident{Df}\langle v_0, \ldots, v_n\rangle$
and a matching kind of expressions. We specify how the \kvd{prev} keyword is translated into a
$\ident{prev}_\cclrd{r}$ operation of the target language and we also add a typing rule
(\emph{df}) that checks the types of the elements of the stream and also guarantees
that the number of elements in the stream matches the number in the coeffect.
The additional reduction rules mirror the semantics that we discussed in
Example~\ref{thm:semantics-indexed-list} when discussing the indexed list comonad.

@@

\paragraph{Properties.}
Now consider a target language consisting of the core (ML-subset) defined by the syntax,
reduction rules and typing rules given in Figure~\ref{fig:semantics-target} and comonadically-inspired
primitives defined in Figure~\ref{fig:semantics-ext} and also concrete notion of comonadically-inspired
value and reduction rules for dataflow as defined in Figure~\ref{fig:semantics-ext-df}.

@@

\paragraph{Parametric effect monads.}
Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our
indexed comonads) are closely related to our definition.  Although presented in a more general
categorical framework (and using monads), the model (i) defines the \ident{unit} operation only on the
unit of a monoid and (ii) the \ident{bind} operation composes effect annotations using the provided
monoidal structure.

@@

\begin{itemize}
\item Semantically, coeffects capture different notions of computation. As demonstrated in
  Chapter~\ref{ch:pathways}, coeffects track additional contextual properties required by a
  computation, many of which cannot be captured by a monad (\eg~liveness or dataflow).
  In terms of program analysis \cite{other-optcomp}, monads capture forward dataflow analyses and
  comonads correspond to backward dataflow analyses.

@@

\noindent
Despite the differences, our implicit parameters resemble, in many ways, the \emph{reader} monad.
As discussed in Section~\ref{sec:semantics-related-monads}, the \emph{reader} monad is semantically
equivalent to the \emph{product} comonad when we consider just sequential composition. For a
language with lambda abstraction, we need a slight extension to the usual treatment of monads in
order to model implicit parameters using a monad.

@@

\noindent
This equivalence shows an intriguing relationship between the \emph{product comonad} and
\emph{reader monad}, but it cannot be extended beyond that. In particular, comonads that model
dataflow computations or liveness do not have a corresponding monadic structure.
This equivalence holds for monads and comonads (as well as \emph{indexed} monads
and comonads), but it does not extend to \emph{flat} indexed comonads which also provide
the $\ident{merge}_{\cclrd{r}, \cclrd{s}}$ operation to model context merging. This can be
supported in monadic computations by adding an additional operation discussed next.

### 2016-05-16 text/Chapter06.tex
Spell checking

\item We give a \emph{translational semantics} (Section~\ref{sec:struct-transl}) that translates
  programs from the structural coeffect calculus into a simple functional language with
  uninterpreted comonadically-inspired primitivies. We give concrete operational semantics for the
  target language for one of our sample languages and show that well-typed programs, produced by
  translation from the coeffect calculus do not get stuck.
\end{itemize}

@@

\paragraph{Subcoeffecting.}
Finbally, Figure~\ref{fig:struct-types} (c) shows the subcoeffecting rule (\emph{sub}).
As with the flat coeffect calculus, we include it mainly to guide the understanding of the
system, but we treat it as an additional rule and do not discuss it in the subsequent theory.
The rule would be needed in languages with conditionals and other language features. It is
worth noting that allowing subcoeffecting on coeffect scalars belonging to individual variables
is sufficient for structural systems.

@@

\paragraph{Lambda abstraction.}
In lambda abstraction (\emph{abs}) in Figure~\ref{fig:struct-types}, we assume that the bound
varible is the last variable of the context. In the syntax-directed system, we do not make the
same assumption. Instead, we use an auxiliary relation $\ident{findvar}_x$ that takes a typing
context $\coctx{\Gamma}{\aclrd{\textbf{t}}}$ and returns a context with the variable $x$ removed
together with its type and coeffect.

@@

\paragraph{Function application.} In the (\emph{app}) rule in Figure~\ref{fig:struct-types}, we
assume that the variable contexts of the two sub-expressions can be merged. This requires that they
contain disjoint variables, which can be always obtained by exchange and contraction. In the
syntax-driven systenm, we merge coeffects of shared variables explicitly. This is done in the
\ident{mergevars} function.

@@

\vspace{0.5em}\noindent\hangindent=0.7cm
Casea (\emph{app}) similarly to (\emph{abs}).
\end{proof}

@@

\begin{remark}[Admisibility of unique typing for implicit parameters]
\label{thm:flat-impl-admis}
  If $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ (using the rules in
  Figure~\ref{fig:struct-syntax-types}) then there is a uniqye typing derivation
  using the typing rules from Figure~\ref{fig:struct-types} with a conclusion
  $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ obtained by induction over
  the original typing derivation as folows:

@@

\paragraph{Expressions.}
A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. The
semantics is written as composition of morphisms using a number of auxiliary definitions. Due to
the eqivalence between Cartesian Closed Categories and the $\lambda$-calculus, we will treat it
as specifying translation to a target functional language in Section~\ref{sec:struct-transl}.

@@

\paragraph{Properties.}
The concrete categortical semantics presented in this section is a generalization of the concrete
semantics given when introducing context-aware programming languages in Chapter~\ref{ch:applications}.

@@

\figcaption{Trnaslation from a structural coeffect calculus}
\label{fig:struct-translation}
\end{figure*}

@@

\begin{theorem}[Well-typedness of the translation]
\label{thm:struct-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$
written in a structural context-aware programming langugae that is translated to the target language as
(we write $\ldots$ for the omitted part of the translation tree):
%
\begin{equation*}
  \semdeff
    {~~(\ldots)~~}
    {\coctx{}{\alift{}} \vdash e : \tau }
    {(\ldots)}
    { ~~f }
\end{equation*}
%
Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{()}{\alift{}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation. Given a judgement
$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\aclrd{\textbf{c}}} \vdash e : \tau$ where
$\aclrd{\textbf{c}} = \alift{\cclrd{c_1},\ldots,\cclrd{c_n}}$, the translation
constructs a function of type $\ctyp{\aclrd{\textbf{c}}}(\sem{\tau_1}\xtimes\ldots\xtimes\sem{\tau_n}}) \rightarrow \sem{\tau}$.

@@

When type-checking expressions that create the \ident{Df} values, we use an auxiliary judgement
$\Gamma \vdash_{\ident{vec}} e : \tau, \cclrd{n}$. The judgement checks that a vector of expressions
$\textbf{e}$ contains exactly $\cclrd{n}$ expressions of type $\tau$. This is caputred by the
(\emph{vec}) rule, which is then used to check individual elements of the context in the
(\emph{df}) rule.

>>>>

\item We give a \emph{translational semantics} (Section~\ref{sec:struct-transl}) that translates
  programs from the structural coeffect calculus into a simple functional language with
  uninterpreted comonadically-inspired primitives. We give concrete operational semantics for the
  target language for one of our sample languages and show that well-typed programs, produced by
  translation from the coeffect calculus do not get stuck.
\end{itemize}

@@

\paragraph{Subcoeffecting.}
Finally, Figure~\ref{fig:struct-types} (c) shows the subcoeffecting rule (\emph{sub}).
As with the flat coeffect calculus, we include it mainly to guide the understanding of the
system, but we treat it as an additional rule and do not discuss it in the subsequent theory.
The rule would be needed in languages with conditionals and other language features. It is
worth noting that allowing subcoeffecting on coeffect scalars belonging to individual variables
is sufficient for structural systems.

@@

\paragraph{Lambda abstraction.}
In lambda abstraction (\emph{abs}) in Figure~\ref{fig:struct-types}, we assume that the bound
variable is the last variable of the context. In the syntax-directed system, we do not make the
same assumption. Instead, we use an auxiliary relation $\ident{findvar}_x$ that takes a typing
context $\coctx{\Gamma}{\aclrd{\textbf{t}}}$ and returns a context with the variable $x$ removed
together with its type and coeffect.

@@

\paragraph{Function application.} In the (\emph{app}) rule in Figure~\ref{fig:struct-types}, we
assume that the variable contexts of the two sub-expressions can be merged. This requires that they
contain disjoint variables, which can be always obtained by exchange and contraction. In the
syntax-driven system, we merge coeffects of shared variables explicitly. This is done in the
\ident{mergevars} function.

@@

\vspace{0.5em}\noindent\hangindent=0.7cm
Case (\emph{app}) similarly to (\emph{abs}).
\end{proof}

@@

\begin{remark}[Admisibility of unique typing for implicit parameters]
\label{thm:flat-impl-admis}
  If $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ (using the rules in
  Figure~\ref{fig:struct-syntax-types}) then there is a unique typing derivation
  using the typing rules from Figure~\ref{fig:struct-types} with a conclusion
  $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ obtained by induction over
  the original typing derivation as follows:

@@

\paragraph{Expressions.}
A semantics of structural coeffect calculi is shown in Figure~\ref{fig:struct-semantics}. The
semantics is written as composition of morphisms using a number of auxiliary definitions. Due to
the equivalence between Cartesian Closed Categories and the $\lambda$-calculus, we will treat it
as specifying translation to a target functional language in Section~\ref{sec:struct-transl}.

@@

\paragraph{Properties.}
The concrete categorical semantics presented in this section is a generalization of the concrete
semantics given when introducing context-aware programming languages in Chapter~\ref{ch:applications}.

@@

\figcaption{Translation from a structural coeffect calculus}
\label{fig:struct-translation}
\end{figure*}

@@

\begin{theorem}[Well-typedness of the translation]
\label{thm:struct-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$
written in a structural context-aware programming language that is translated to the target language as
(we write $\ldots$ for the omitted part of the translation tree):
%
\begin{equation*}
  \semdeff
    {~~(\ldots)~~}
    {\coctx{}{\alift{}} \vdash e : \tau }
    {(\ldots)}
    { ~~f }
\end{equation*}
%
Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{()}{\alift{}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation. Given a judgement
$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\aclrd{\textbf{c}}} \vdash e : \tau$ where
$\aclrd{\textbf{c}} = \alift{\cclrd{c_1},\ldots,\cclrd{c_n}}$, the translation
constructs a function of type $\ctyp{\aclrd{\textbf{c}}}(\sem{\tau_1}\xtimes\ldots\xtimes\sem{\tau_n}}) \rightarrow \sem{\tau}$.

@@

When type-checking expressions that create the \ident{Df} values, we use an auxiliary judgement
$\Gamma \vdash_{\ident{vec}} e : \tau, \cclrd{n}$. The judgement checks that a vector of expressions
$\textbf{e}$ contains exactly $\cclrd{n}$ expressions of type $\tau$. This is captured by the
(\emph{vec}) rule, which is then used to check individual elements of the context in the
(\emph{df}) rule.

### 2016-05-16 text/Chapter07.tex
Spell checking

The implementation directly follows the thoery presented in Chapters~\ref{ch:flat}, \ref{ch:semantics}
and \ref{ch:structural}. It consists of a common framework that provides type checking and
translation to a simple functional target language with comonadically-inspired primitives. Each
concrete context-aware language then adds a domain-specific rule for choosing a unique typing
derivation (as discussed in Section~\ref{sec:flat-unique}) together with a domain-specific
definition of the comonadically-inspired primitives that define the runtime semantics
(see Section~\ref{sec:semantics-proofs}).

The main goal of the implementation is to show that the theory is practically useful and to present it
in a more practical way. However, we do not intend to build a complete real-world programming language.
For this reason, the implementation is available primarilly as an interactive essay\footnote{The
\emph{interactive essay} format is based on Bret Victor's work on \emph{explorable explanations} \cite{essay-explorable}
and is further explained in Section~\ref{sec:impl-essay-features}} online at \url{http://tomasp.net/coeffects}.

@@

\begin{itemize}
\item We discuss how the implementation follows the theory (Section~\ref{sec:impl-theory}) presented earlier.
  This applies to the implementation of the \emph{type checker} and the implementation of the \emph{translation}
  to a simple target language that is then iterpereted. We also discuss how the common framework makes it
  easy to implement additional context-aware languages (Section~\ref{sec:impl-theory-ext}).

@@

The theory discussed so far provides the two key components of the implementation. In
Chapter~\ref{ch:flat}, we discussed the type checking of context-aware programs and
Chapter~\ref{ch:semantics} models the execution of context-aware programs (in terms of
translation and operational semantics). For structural coeffects, the same components are discussed
in Chapter~\ref{ch:structural}. In this section, we discuss how teose provide foundation for the
implementation.

@@

\begin{enumerate}
\item A domain-specific function abstraction rule that resolves the ambiguity in the general
  (\emph{abs}) rule of the flat coeffect calculus. For liveness, the handling would be the same
  as for atafloww, but for other flat coeffect systems, another resolution mechanism might be used
  instead.

@@

We first consider two case studies of how coeffect type checking works. The first one exposes the
ambiguity resolution algorithm for the typing of implicit parameters and the second one exposes the
difference between flat and structural system forataflowow.

@@

\paragraph{Interactive program excution.}

@@

The interactive program execution lets the reader run sample programs, but not in a particuarly realistic
context. To show a more real-world scenario, the essay includes a widget shown in Figure~\ref{fig:essay-df}.
This lets the user write a function taking a stream of X and Y coordinates and calculate value
based on the current and past values of the mouse pointer. The X and Y values, together with the
result are plotted using a live chart.

@@

First, it merges all variables into a single (comonadic) value representing
the context. Second, there is a flat and structural variation of the system. For these two reasons,
understanding the translation based just on the rules is harder than for monads. The essay lets
readers see translation for carefuly chosen case studies that illustrate the key aspects of the
translation (some discussed in Section~\ref{sec:impl-case-transl}), but also for their own
programs encouraging \emph{active reading}. Sample input and output are shown in
Figure~\ref{fig:essay-transl}.

@@

\item Type-checked programs in context-aware languages are translated to simple target
  functional subset of the language in \strf{translation.fs}; \strf{evaluation.fs} then interprets
  programs in the target language. The interpretation does not handle the source
  language and so programs containing context-aware contstructs cannot be interpreted directly.

>>>>

The implementation directly follows the theory presented in Chapters~\ref{ch:flat}, \ref{ch:semantics}
and \ref{ch:structural}. It consists of a common framework that provides type checking and
translation to a simple functional target language with comonadically-inspired primitives. Each
concrete context-aware language then adds a domain-specific rule for choosing a unique typing
derivation (as discussed in Section~\ref{sec:flat-unique}) together with a domain-specific
definition of the comonadically-inspired primitives that define the runtime semantics
(see Section~\ref{sec:semantics-proofs}).

The main goal of the implementation is to show that the theory is practically useful and to present it
in a more practical way. However, we do not intend to build a complete real-world programming language.
For this reason, the implementation is available primarily as an interactive essay\footnote{The
\emph{interactive essay} format is based on Bret Victor's work on \emph{explorable explanations} \cite{essay-explorable}
and is further explained in Section~\ref{sec:impl-essay-features}} online at \url{http://tomasp.net/coeffects}.

@@

\begin{itemize}
\item We discuss how the implementation follows the theory (Section~\ref{sec:impl-theory}) presented earlier.
  This applies to the implementation of the \emph{type checker} and the implementation of the \emph{translation}
  to a simple target language that is then interpreted. We also discuss how the common framework makes it
  easy to implement additional context-aware languages (Section~\ref{sec:impl-theory-ext}).

@@

The theory discussed so far provides the two key components of the implementation. In
Chapter~\ref{ch:flat}, we discussed the type checking of context-aware programs and
Chapter~\ref{ch:semantics} models the execution of context-aware programs (in terms of
translation and operational semantics). For structural coeffects, the same components are discussed
in Chapter~\ref{ch:structural}. In this section, we discuss how these provide foundation for the
implementation.

@@

\begin{enumerate}
\item A domain-specific function abstraction rule that resolves the ambiguity in the general
  (\emph{abs}) rule of the flat coeffect calculus. For liveness, the handling would be the same
  as for dataflow, but for other flat coeffect systems, another resolution mechanism might be used
  instead.

@@

We first consider two case studies of how coeffect type checking works. The first one exposes the
ambiguity resolution algorithm for the typing of implicit parameters and the second one exposes the
difference between flat and structural system for dataflow.

@@

\paragraph{Interactive program execution.}

@@

The interactive program execution lets the reader run sample programs, but not in a particularly realistic
context. To show a more real-world scenario, the essay includes a widget shown in Figure~\ref{fig:essay-df}.
This lets the user write a function taking a stream of X and Y coordinates and calculate value
based on the current and past values of the mouse pointer. The X and Y values, together with the
result are plotted using a live chart.

@@

First, it merges all variables into a single (comonadic) value representing
the context. Second, there is a flat and structural variation of the system. For these two reasons,
understanding the translation based just on the rules is harder than for monads. The essay lets
readers see translation for carefully chosen case studies that illustrate the key aspects of the
translation (some discussed in Section~\ref{sec:impl-case-transl}), but also for their own
programs encouraging \emph{active reading}. Sample input and output are shown in
Figure~\ref{fig:essay-transl}.

@@

\item Type-checked programs in context-aware languages are translated to simple target
  functional subset of the language in \strf{translation.fs}; \strf{evaluation.fs} then interprets
  programs in the target language. The interpretation does not handle the source
  language and so programs containing context-aware constructs cannot be interpreted directly.
